Robustness of MinD oscillation in \emph{Escherichia coli} with diverse cell shapes

### 2014-02-05 paper/paper.tex
updates to methods and pill section of paper Tested everything on node32.

Our diffusion and reaction rates are shown below.  We are interested
primarily in the effect of cellular size and shape on the protein
oscillations, so we follow Huang\cite{huang2003dynamic} and do not
deviate from the wild type values used in the cited work (see values
below).  Huang's simulations use total MinD and MinE concentrations of
$1,000/\mu m$ and $350/\mu m$, respectively, in a cylindrical cell of
radius $0.5\mu m$, and in our (non-cylindrical) cells we use the same
number of proteins per unit volume.  These concentration values are
1273 $\mu m^{-3}$ and 445 $\mu m^{-3}$, respectively.
\begin{gather*}
  \mathcal{D}_D = \mathcal{D}_{E} = 2.5\micron^2/\text{sec}\\
  k_D^{\textrm{ADP $\rightarrow$ ATP}} = 1/\textrm{sec,  }
  k_D = 0.025 \micron /\textrm{sec}\\
  k_{dD} = 0.0015 \micron^3/ \textrm{sec,  }
  k_{de} = 0.7/\textrm{sec}\\
  k_E = 0.093 \micron^3 /\textrm{sec}.
\end{gather*}
A 3D grid is then constructed in cartesian coordinates, with a grid
spacing of .05\micron. We define our cell shapes and solve the
reaction-diffusion partial differential equations numerically to
observe the time evolution of the MinD and MinE concentrations inside
the cell.

@@

We next see the formation of MinE `rings' on the membranes of the
polar caps.  As the freely diffusing MinE enters the cap portion it
begins to encounter concentrated membrane bound MinD.  The MinE binds
to this MinD and remains for a time on the walls in compound form.
When it is released, it has the two options of diffusing naturally
outward or inward towards the cell.  If it diffuses inward, there is
less cross sectional wall for it to bump into and also less
concentrated MinD:ATP to react with.  If it diffuses outward, the
curvature of the cell wall cuases there to be a greater probability
per unit time that it will encounter the cell membrane (for the same
reason outlined for the MinD:ATP above) where there is more membrane
bound MinD:ATP. If it does diffuse in this directin, it will likely
react with MinD:ATP before getting very far into the encap.  This
process results in the observed qualitative 'encroaching MinE ring'
behavoir that releases MinD:ADP back into the cytoplasm.  This
MinD:ADP will diffuse for a time before spontanosly undergowing
nucleotide exchange and reaching the MinD:ATP state that
can bind to the membrane again.  There is a strong chance that it has
by this time diffused out of the polar region and is on its way to the
opposite polar region of the cell.  If it hasn't then it will reattach
to the MinD:ATP on the cell wall deeper into the endcap and will
eventually be converged upon by the MinE ring, only to be released
again in MinD:ADP form.

@@

We mean to investigate the geometric limits of the Min system
oscillations, as observed by Mannik \emph{et
  al.}~\cite{mannik2012robustness}, so we model the Min system in a
variety of cell shapes.  We begin with naturally occuring pill-shaped
cells and then proceed to examine a number of flattened pancake-like
shapes, which reflect the experiment of Mannik \emph{et al.}, in which
bacteria were confined within a thin slit~\cite{mannik2012robustness}.
Our pill shapes differ from those of Huang \emph{et al.} in that they
are cylinders with hemisphere endcaps instead of pure cylindrical
shapes.  Our cylinderidrical radius is $0.5\micron$ and the lengths
of our cells (measured between the tips of the endcaps) are
$5\micron$, $4\micron$, $3\micron$, and $2.5\micron$.

With our flattened cell shapes we mean to study cells that are
parrallel to those lodged into crevices by Mannik so we give them
height of $0.25\micron$ \cite{mannik2012robustness}.  Viewed from the top
down the cells will have the shapes described below and viewed from
the side they have at their edges a semicircular protrusion (one may
imagine the edges of a pancake).

Our first group of (three) flattened cell shapes are triangular when
viewed from above.  One of them is equilateral, one has a 3-4-5 right
angle shape, and the last is iscoscolese.

Our second group of (two) flattened cells are designed to replicate
those published by Mannik, in order to compare the results of the
simulation model with the experimental results in that work.

Our third group of (three) flattened cells are our own creations,
designed to investigate different patterns of protein oscillation
behavoir.

Kubitschek has shown in multiple experiments that at the time of cell
division cells have a volume that is within a range of roughly
$1\micron^3$ to $2\micron^3$~\cite{kubitschek1990cell,
  kubitschek1968linear}.  We follow Huang's
simulations\cite{huang2003dynamic} and Mannik's experiments and model
cells that are slightly larger than this range.  Amoung our flattened
cells the two dimensional length scales are tuned so that total cell
volumes are very close to $3\micron^3$. \jeffsays{I'm not sure how
  true this is.  We should go through and check.  There've been a few
  implementations of printouts reporting volume in the cpp}

@@

At $t=0$ in Figs.~\ref{image-p}~and~\ref{box-p}, the cytoplasmic MinE
is concentrated in the lower portion of the cell, and is diffusing
upward.  The MinD protein is initially concentrated on the membrane in
the upper portion of the cell.  At the boundary between the
cytoplasmic MinE and the membrane-bound MinD, there is a ring of the
membrane-bound MinE:MinD complex, which is moving towards the upper
end of the cell, removing the MinD from the membrane as it goes.  The
MinD is removed from the membrane in its MinD:ADP form, which begins
to accumulate at the upper end of the cell, peaking peaking at around
10 seconds.

@@

\begin{figure}
  \includegraphics[width=\columnwidth]{../data/shape-p/plots/box-plot_D--p-300-50-0-0-1500}
  \caption{Total protein fluctuation in the right, middle, and left
    parts of a natural pill shaped cell.  The cell is the same as is
    shown in Fig.~\ref{image-p}.  The time axis is also identical to
    that shown in Fig.~\ref{image-p}, and shows one period of the
    oscillation after the system has settled down.  The vertical axis
    shows stacked the total number of proteins that are of four
    different compound stages and in the right section of the cell
    (bottom four colors), in the middle of the cell (middle four
    colors) and in the left section of the cell (top four colors). The
    different forms of protein changes while the total number of MinD
    protein in the cell remains constant, shown by the fact that the
    very top line is horizontal.}
  \label{box-p}
\end{figure}

While Fig.~\ref{image-p} is helpful for visualizing
the process occuring in the Min system, it is does not show the
relative number of proteins in each state.  To visualize this
information, we plot in Fig.~\ref{box-p} the fraction of MinD found in
each stage of the cycle in each of three sections of the cell---top,
middle and bottom---as a function of time over one period.  The time
axis is the same as in Fig.~\ref{image-p}.  Looking at the evolution
of the latter, we see an initial high concentration of cytoplasmic
MinD:ATP, as the this protein diffuses into the polar region, followed
by a dip in this and large spike in membrane bound MinD:ATP, as the
protein clusters onto the polar cell wall, followed by a spike in the
memebrane bound MinE:MinD:ATP stage of protein, together with a slight
rise in cytoplasmic MinD:ADP as the MinD dettaches from the wall.

results for pill shaped cell of length
$5\mu$ and radius $.5\mu$.  The plot shows both the transerence
between different stages of protein compound and movement within the
cell.  Starting at the beginning of a period, we can see that first
there is a spike on the left of MinD-ATP attached to the wall that is
accompanied by a growing number of cytoplasmic MinD-ATP and MinD-ADP
that are entering the left section, followed by a high peak in
MinD-MinE-ATP.  This peak gives way through the middle to peaks on the
right side of the cell, the first being a small bump in cytoplasmic
MinD-ATP that dissapears into a much larger spike in wall attached
MinD-ATP, which is followed by a peak in wall attached MinD-MinE-ATP.
The pattern during each period clearly shows the cycle of proteins
rections, as well as the very regular nature of the pole to pole
protein oscillations.

We can see a
clear difference in period between the $5\mu$, $4\mu$, and $2\mu$
length cells.  The oscillation periods for our $4\mu$ show excellent
agreement with the simulations of Huang et all
\cite{huang2003dynamic}.  This is to be expected since we begin our
simulations with their reported wild type concentrations of MinD and
MinE proteins. The periods for the different pill lengths are shown in
the table below.  We can see a direct correltation between size and
period, as the proteins must diffuse further in a longer cell before
they accumulate on the opposite wall.

\begin{figure}
  \includegraphics[width=\columnwidth]{../data/shape-p/plots/ave-plot_D--p-300-50-0-0-1500}
  \caption{Average protein density per unit area of wall for the wall
    attached stages of the protein cycle.  Shown in the right, middle,
    and left parts of a 5$\micron$ by 1$\micron$ pill shaped cell (above)
    and a 3$\micron$ by 1$\micron$ pill shaped cell (below).}
  \label{ave-per-area-plot-pill}
\end{figure}

Figure \ref{ave-per-area-plot-pill} shows the ave protein per unit
area that is wall attached in the middle, right, and left sections of
the pill shaped cell.  This is an important test of the central idea
behind the utility of the Min protein system.  The system is meant to
place Min proteins on the walls in heavier concentrations in the caps
of the cell and lower concentrations in the middle section, so that
the FtZ polymer may build up along the center walls.  We can see from
the plot of the 3$\micron$ by 1$\micron$ cell that the MinD proteins reach
concentration peaks in the caps that are higher than peaks in the
center.  \fixme{is the next part true?} Integrals of these plotted values show that the time averaged
concentrations are likewise higher in the caps than in the center.

\subsection{Triangular Shapes}
\begin{figure}
  \includegraphics[width=\columnwidth]{../data/shape-triangle/plots/box-plot_D--triangle-25-400-400-400-1500}
  \includegraphics[width=\columnwidth]{../data/shape-triangle/plots/box-plot_D--triangle-25-500-300-500-1500}
  \includegraphics[width=\columnwidth]{../data/shape-triangle/plots/box-plot_D--triangle-25-600-480-360-1500}
  \caption{Total protein fluctuation in the right, middle, and left
    parts of an equilateral (top), iscoscolese (middle), and 3-4-5
    (bottom) triangle.  The vertical axis shows stacked the total
    number of proteins that are of four different compound stages and
    in the different sections of the cell.}
  \label{box-triangle}
\end{figure}

@@

Fifteen

>>>>

Our diffusion and reaction rates are shown below.  We are interested
primarily in the effect of cellular size and shape on the protein
oscillations, so we follow Huang\cite{huang2003dynamic} and do not
deviate from the wild type values used in the cited work (see values
below).  Huang's simulations use total MinD and MinE concentrations of
$1,000/\mu m$ and $350/\mu m$, respectively, in a cylindrical cell of
radius $0.5\mu m$, and in our (non-cylindrical) cells we use the same
number of proteins per unit volume.  These concentration values are
1273 $\mu m^{-3}$ and 446 $\mu m^{-3}$, respectively.
\begin{gather*}
  \mathcal{D}_D = \mathcal{D}_{E} = 2.5\micron^2/\text{sec}\\
  k_D^{\textrm{ADP $\rightarrow$ ATP}} = 1/\textrm{sec,  }
  k_D = 0.025 \micron /\textrm{sec}\\
  k_{dD} = 0.0015 \micron^3/ \textrm{sec,  }
  k_{de} = 0.7/\textrm{sec}\\
  k_E = 0.093 \micron^3 /\textrm{sec}.
\end{gather*}
A 3D grid is then constructed in cartesian coordinates, with a grid
spacing of .05\micron. We define our cell shapes and solve the
reaction-diffusion partial differential equations numerically to
observe the time evolution of the MinD and MinE concentrations inside
the cell.

We mean to investigate the geometric limits of the Min system
oscillations as observed by Mannik \emph{et
  al.}~\cite{mannik2012robustness}, so we model the Min system in a
variety of cell shapes and sizes.  Here we present a selection of
those, first naturally occuring pill-shaped cells and then a
number of flattened out shapes which reflect the experiments of Mannik
\emph{et al.}, in which bacteria are confined within a thin slit of
height $0.25\micron$ ~\cite{mannik2012robustness}. Viewed from the top
down the cells will have the shapes described below and viewed from
the side they have at their edges a semicircular protrusion (one may
imagine the edges of a pancake). Central to our observations are the
effects of cell size and so we show expanded cells that have the same
thickness but increased dimensions in the two planar directions.  We
present two shapes that are designed to replicate those published by
Mannik, equilateral and iscosolese triangle shapes, and a 'V' shape of
our own creation in order to best illustrate the effect of cell size
on oscillation stability.

%% Our pill shapes differ from those of Huang \emph{et al.} in that they
%% are cylinders with hemisphere endcaps instead of pure cylindrical
%% shapes.  Our cylinderidrical radius is $0.5\micron$ and the lengths
%% of our cells (measured between the tips of the endcaps) are
%% $5\micron$, $4\micron$, $3\micron$, and $2.5\micron$.

Kubitschek has shown in multiple experiments that at the time of cell
division cells have a volume that is within a range of roughly
$1\micron^3$ to $2\micron^3$~\cite{kubitschek1990cell,
  kubitschek1968linear}.  We follow Huang's
simulations\cite{huang2003dynamic} and Mannik's experiments and model
cells that are slightly larger than this range.

@@

We next see the formation of MinE `rings' on the membranes of the
polar caps.  As the freely diffusing MinE enters the cap portion it
begins to encounter concentrated membrane bound MinD.  The MinE binds
to this MinD and remains for a time on the walls in compound form.
When released, if it diffuses inward, there is less cross sectional
wall for it to bump into and also less concentrated MinD:ATP to react
with, whereas if it diffuses outward the opposite is the case.  This
process results in the observed qualitative 'encroaching MinE ring'
behavoir that releases MinD:ADP back into the cytoplasm.  This
MinD:ADP will diffuse for a time before spontanosly undergowing
nucleotide exchange and reaching the MinD:ATP state that can bind to
the membrane again.  There is a strong chance that it has by this time
diffused out of the polar region and is on its way to the opposite
polar region of the cell.  If it hasn't then it will reattach to the
MinD:ATP on the cell wall deeper into the endcap and will eventually
be converged upon by the MinE ring, only to be released again in
MinD:ADP form.

@@

@@

At $t=0$ in Figs.~\ref{image-p}, the cytoplasmic MinE
is concentrated in the lower portion of the cell, and is diffusing
upward.  The MinD protein is initially concentrated on the membrane in
the upper portion of the cell.  At the boundary between the
cytoplasmic MinE and the membrane-bound MinD, there is a ring of the
membrane-bound MinE:MinD complex, which is moving towards the upper
end of the cell, removing the MinD from the membrane as it goes.  The
MinD is removed from the membrane in its MinD:ADP form, which begins
to accumulate at the upper end of the cell, peaking peaking at around
10 seconds.

@@

%% \begin{figure}
%%   \includegraphics[width=\columnwidth]{../data/shape-p/plots/box-plot_D--p-300-50-0-0-1500}
%%   \caption{Total protein fluctuation in the right, middle, and left
%%     parts of a natural pill shaped cell.  The cell is the same as is
%%     shown in Fig.~\ref{image-p}.  The time axis is also identical to
%%     that shown in Fig.~\ref{image-p}, and shows one period of the
%%     oscillation after the system has settled down.  The vertical axis
%%     shows stacked the total number of proteins that are of four
%%     different compound stages and in the right section of the cell
%%     (bottom four colors), in the middle of the cell (middle four
%%     colors) and in the left section of the cell (top four colors). The
%%     different forms of protein changes while the total number of MinD
%%     protein in the cell remains constant, shown by the fact that the
%%     very top line is horizontal.}
%%   \label{box-p}
%% \end{figure}

\subsection{Triangular Shapes}
\begin{figure}
  %%\includegraphics[width=\columnwidth]{../data/shape-triangle/plots/box-plot_D--triangle-25-400-400-400-1500}
  %%\includegraphics[width=\columnwidth]{../data/shape-triangle/plots/box-plot_D--triangle-25-500-300-500-1500}
  %%\includegraphics[width=\columnwidth]{../data/shape-triangle/plots/box-plot_D--triangle-25-600-480-360-1500}
  \caption{Total protein fluctuation in the right, middle, and left
    parts of an equilateral (top), iscoscolese (middle), and 3-4-5
    (bottom) triangle.  The vertical axis shows stacked the total
    number of proteins that are of four different compound stages and
    in the different sections of the cell.}
  \label{box-triangle}
\end{figure}

@@
